Gelation as condensation frustrated by hydrodynamics and mechanical tension

### 2015-10-21 manuscript4.tex
no sealing of the end of the channel 10 mm2 x 200 um for the cell

To access roles of mechanics in the late stage, we realize a density matching of the order of 10$^{-4}$ (gravitational Peclet number $Pe< 10^{-7}$) and use a very accurate particle tracking method \cite{leocmach2013novel}. 

@@

We enclose the suspension in a thin microscopy cell sketched in Fig.~\ref{fig:cell_vs_cap}a. The bottom wall of the cell is an osmotic membrane providing contact with a long channel full of the same solvent mixture. We introduce solid salt at the other end of the channel that we seal immediately. Salt dissolution and subsequent migration of the ions along the channel and through the membrane induce screening of the electrostatic repulsion, well revealing the depletion potential due to the pre-existing polymers. The time needed for the ions to diffuse from the membrane across the cell thickness is of the same order of magnitude as the Brownian time of the particles $\tau_B=10$ s. This method causes uniform gelation starting from the homogeneous Wigner crystal state without any solvent flow, allowing \textit{in situ} confocal microscopy observation throughout the process from $t \sim 0$.

@@

\vspace{1cm}
\noindent
{\bf METHODS}

\vspace{0.5cm}
\noindent
{\bf Experimental.}

@@

The colloids and polymers are contained in an observation cell ($1$ mm$^2$ $\times$ 200 $\mu$m) made of glass in contact with an half-open glass channel approximately 400 times larger in volume, via a millipore filter with pore size of 100 nm that allows the salt through but neither polymer nor colloid (see Fig.~\ref{fig:cell_vs_cap}a). The channel is filled with the same solvents at density matching composition. At the beginning of the experiment, solid tetrabutylammonium bromide (Fluka) is introduced to the channel which is immediately sealed to prevent evaporation. Data acquisition starts within 30 s after salt introduction. Our procedure induces practically no solvent flow in the observation cell. We confirmed the presence of undissolved salt several days after mixing, indicating that the observation cell was brought to saturation concentration.

Given the diffusion constants of Bromide and alkyl cation ($6$ and $2 \times 10^{-10}$ m$^2$/s~\cite{Campbell2005}), we estimate the characteristic diffusion time of salt from top to bottom of the order of 10 s. Therefore, we reach uniform final salt concentration into the observation cell within only a few Brownian times of the colloids. Indeed we measured a delay of about 1 min between the aggregation at the bottom and at the top of the cell. We define the initial time of the aggregation process when the maximum of the $g(r)$ jumps from the lattice constant of the Wigner crystal to the hard-core diameter $\sigma$.

We collect the data on a Leica SP5 confocal microscope, using 532 nm laser excitation. The temperature was controlled on both stage and objective lens, allowing a more precise density matching. The scanned volume is at least $82 \times 82 \times 85$ $\mu$m$^3$. The particle coordinates are tracked in three dimensions (3D) with an accuracy of around $0.03\sigma$~\cite{LeocmachColloids}.

\vspace{0.5cm}
\noindent
{\bf Analysis.}

From direct confocal measurements~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=2.75$ $\mu$m) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Spinodal line on Fig.~\ref{fig:phasediag}b are calculated from this size ratio using the generalized free volume theory~\cite{Fleer2008}.

In principle the attraction well of the depletion extends to $\sigma+2R_g$, however, resolution-dependent tracking imprecisions and systematic errors do not give a precise enough estimate of such short distance. Therefore we consider two particles bonded when their distance is shorter the first minimum of $g(r)$, i.e. 3.55 $\mu$m. This defines the bond graph that we analyse using connected components and shortest path analysis algorithms~\cite{Hagberg2008}.

@@

\clearpage 
\begin{figure}
%\includegraphics[width=10cm]{figure2.pdf}
\includegraphics{figs/phasediag.pdf}
\caption{\textbf{Different regimes of gelation.} \textbf{a,} Phase diagram. Experimental points are categorised based on the final state obtained in reservoir cell. The spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of Fig.~\ref{fig:cell_vs_cap}b-c is highlighted by a square. \textbf{b,} Comparison of system evolution in terms of largest cluster extent and of mean coordination number. By increasing density: $\phi=4.2,\,8,\,16,\,27~\%$, $c_p=1,\,1.5,\,1.2,\, 1$ g/$\ell$ for 
%\textcolor{red!40!yellow}
{$\blacklozenge$}, 
%\textcolor{red!80!yellow}
{$\circ$}, 
%\textcolor{red!80!black}
{\tiny$\blacksquare$} and 
%\textcolor{red!40!black}
{$\bullet$} respectively. \textbf{c,} Growth of the characteristic wave number for the same samples.}
\label{fig:phasediag}
\end{figure}

>>>>

To access roles of mechanics in the late stage, we realize a density matching of the order of 10$^{-4}$ (gravitational Peclet number $Pe< 10^{-6}$) and use a very accurate particle tracking method \cite{leocmach2013novel}. 

@@

We enclose the suspension in a thin microscopy cell sketched in Fig.~\ref{fig:cell_vs_cap}a. The bottom wall of the cell is an osmotic membrane providing contact with a long channel full of the same solvent mixture. We introduce solid salt at the other end of the channel. Salt dissolution and subsequent migration of the ions along the channel and through the membrane induce screening of the electrostatic repulsion, well revealing the depletion potential due to the pre-existing polymers. The time needed for the ions to diffuse from the membrane across the cell thickness is of the same order of magnitude as the Brownian time of the particles $\tau_B=10$ s. This method causes uniform gelation starting from the homogeneous Wigner crystal state without any solvent flow, allowing \textit{in situ} confocal microscopy observation throughout the process from $t \sim 0$.

@@

\section*{METHODS}

\subsection*{Experimental.}

@@

The colloids and polymers are contained in an observation cell ($10$ mm$^2$ $\times$ 200 $\mu$m) made of glass in contact with an half-open glass channel approximately 400 times larger in volume, via a millipore filter with pore size of 100 nm that allows the salt through but neither polymer nor colloid (see Fig.~\ref{fig:cell_vs_cap}a). The channel is filled with the same solvents at density matching composition. At the beginning of the experiment, solid tetrabutylammonium bromide (Fluka) is introduced to the channel. Data acquisition starts within 30 s after salt introduction. Our procedure induces practically no solvent flow in the observation cell. We confirmed the presence of undissolved salt several days after mixing, indicating that the observation cell was brought to saturation concentration.

Given the diffusion constants of Bromide and alkyl cation ($6$ and $2 \times 10^{-10}$~m$^2$/s~\cite{Campbell2005}), we estimate the characteristic diffusion time of salt from top to bottom of the order of 10~s. Therefore, we reach uniform final salt concentration into the observation cell within only a few Brownian times of the colloids. Indeed we measured a delay of about 1 min between the aggregation at the bottom and at the top of the cell. We define the initial time of the aggregation process when the maximum of the $g(r)$ jumps from the lattice constant of the Wigner crystal to the hard-core diameter $\sigma$.

We collect the data on a Leica SP5 confocal microscope, using 532 nm laser excitation. The temperature was controlled on both stage and objective lens, allowing a more precise density matching. The scanned volume is at least $82 \times 82 \times 85$~$\mu$m$^3$. The particle coordinates are tracked in three dimensions (3D) with an accuracy of around $0.03\sigma$~\cite{LeocmachColloids}.

\subsection*{Analysis.}

From direct confocal measurements~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=2.75$~$\mu$m) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Spinodal line on Fig.~\ref{fig:phasediag}b are calculated from this size ratio using the generalized free volume theory~\cite{Fleer2008}.

In principle the attraction well of the depletion extends to $\sigma+2R_g$, however, resolution-dependent tracking imprecisions and systematic errors do not give a precise enough estimate of such short distance. Therefore we consider two particles bonded when their distance is shorter the first minimum of $g(r)$, i.e. 3.55~$\mu$m. This defines the bond graph that we analyse using connected components and shortest path analysis algorithms~\cite{Hagberg2008}.

@@

\clearpage 
\begin{figure}
%\includegraphics[width=10cm]{figure2.pdf}
\includegraphics{figs/phasediag.pdf}
\caption{\textbf{Different regimes of gelation.} \textbf{a,} Phase diagram. Experimental points are categorised based on the final state obtained in reservoir cell. The spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of Fig.~\ref{fig:cell_vs_cap}b-c is highlighted by a square. \textbf{b,} Comparison of system evolution in terms of largest cluster extent and of mean coordination number. By increasing density: $\phi=4.2,\,8,\,16,\,27~\%$, $c_p=1,\,1.5,\,1.2,\, 1$ g/$\ell$ for 
%\textcolor{red!40!yellow}
{$\blacklozenge$}, 
%\textcolor{red!80!yellow}
{$\circ$}, 
%\textcolor{red!80!black}
{\tiny$\blacksquare$} and 
%\textcolor{red!40!black}
{$\bullet$} respectively. \textbf{c,} Growth of the characteristic wave number for the same samples. 
The line is a scaling law for the intermediate coarsening regime for network-forming phase separation.}
\label{fig:phasediag}
\end{figure}
